Regression Discontinuity Design to Measure Incumbency Advantage in U.S. Congressional Elections

### 2023-04-25 paper.qmd
try cross ref

---
title: "Regression Discontinuity Design to Measure Incumbency Advantage in US Congressional Elections"
subtitle: "Final Project for STAT 530"
author: "Tyler Bagwell and Prayag Gordy"
date: "April 24, 2023"
abstract: "The incumbency effect in U.S. elections is a subject of decades of political science research. Why is it, political scientists ask, that legislative and executive bodies with such low overall approval ratings have the vast majority of its members returning term after term? In this paper, we will use a regression discontinuity design to examine the effects of party incumbency in tightly contested congressional districts. Specifically, we examine the effect of the Democratic margin in an election on the Democratic margin in the following election. We found a local average treatment effect of 17.3 percentage points. Our results further support the literature claiming a large incumbency effect, though we show its strength may be decreasing."
format: pdf
toc: true
number-sections: true
execute:
  echo: false
  message: false
  warning: false
  results: false
---

@@

Incumbents in the U.S. House of Representatives are reelected at astonishing rates. Since 1964, the reelection rate has never dropped below 85%, while the approval rating of the House as a whole has generally remained around 40%, dipping to 20% in recent years. Thus, many political scientists spend their careers researching the advantages of incumbency (Gelman and King, 1990).

Many experts have criticized the apparently strong effects of incumbency. Congressmen can use government funds to visit areas of their district, meet with important political players, and send mail to constituents touting their achievements, all of which indirectly benefit their reelection campaign. Challengers enjoy none of these benefits.

Previous research has found strong effects of incumbency. Candidates running for reelection routinely raise far more money. Further, they may be able to deter challengers simply by being an incumbent. Overall, Cummings found that in similar districts, incumbent congressmen are more likely to win elections than non-incumbent congressmen from the same party.

@@

In practice, it is impossible to determine if the continuity assumption holds for a particular experimental design unless the RDD practitioner has full knowledge of the conditional population densities of the covariates $\mathbf{X}$ with respect to the assigning variable $W$, i.e., $f(\mathbf{X}|W)$, for all covariates that are relevant to $Y$. In empirical RDD, the continuity assumption can be tested with the observed data by (1) testing for the lack of discontinuity in the density of the assignment variable $W$ at threshold $W=w_0$ (McCrary density test) and (2) testing for discontinuity in the first moment of the covariates $\mathbf{X}$ conditional on $W$ at $W=w_0$. This second test is usually performed by using local linear regressions in the proximity of the line $W=w_0$. In less rigorous language, the first test is meant to determine that the odds of treatment assignment of the barely-controls and the barely-treated are purely random in the proximity of the cutoff, while the second test is to determine if the pretreatment characteristics of barely-controls and the barely-treated are indeed very similar.

@@

The data used in this study came from two primary data sets housed by the Harvard Dataverse and the MIT Election Data and Science Lab. The first data set, _U.S. House 1976–2020 Elections_, is a collection of results of US Congressional House elections between 1976 and 2020 for all congressional districts. The primary variables used in this data set were _candidate_, _candidate party_, _candidate votes_, and _total votes for election_. The second data set, _Historical Congressional Legislation and District Demographics 1972-2014_, contains a subset of the pretreatment covariates per congressional district we used for this study, namely, _unemployment rate_, _race_, and _level of education_.

@@

First, we test the continuity of the Democratic vote margin for all $t-1$ elections in our data set. The results of the McCrary density test on the Democratic vote margin at the zero margin cutoff indicates a continuous density on both sides of the cutoff value ($\alpha = 0.05$). A plot of the McCrary density test’s density estimation is provided below.

```{r}
#| fig-cap: McCrary density test and estimation for all $t-1$ Democratic vote margins.

@@

With these three pretreatment covariates chosen, we estimate the discontinuities in their conditional densities conditional on the $t-1$ Democratic election margin at the zero margin value using local linear regression and find that there are no statistically significant discontinuities in either three covariates’ densities. This result can also be expected just by visual inspection of the covariate value versus the $t-1$ Democratic election margin.

```{r}
#| fig-cap: Three pretreatment covariates versus $t-1$ Democratic vote margin.

@@

We found a clear discontinuity from a slight Democratic loss to a slight Democratic victory. A district that votes barely against the Democrat in election $t-1$ tends to shift further away from Democrats in election $t$, while districts that only marginally elect a Democrat in election $t-1$ move toward Democrats in election $t$. This is a clear effect of the treatment — in this case, electing a Democrat — on the Democratic margin in the next election.

```{r}
#| fig-cap: RDD analysis of 4,800 elections from 1976 through 2020. The local average treatment effect is 0.173.

@@

@@

backwards_rdd(DF)
```

@@

>>>>

---
title: "Regression Discontinuity Design to Measure Incumbency Advantage in U.S. Congressional Elections"
subtitle: "Final Project for STAT 530, Spring 2023"
author: "Tyler Bagwell and Prayag Gordy"
abstract: "The incumbency effect in U.S. elections is a subject of decades of political science research. Why is it, political scientists ask, that legislative and executive bodies with such low overall approval ratings have the vast majority of its members returning term after term? In this paper, we will use a regression discontinuity design to examine the effects of party incumbency in tightly contested congressional districts. Specifically, we examine the effect of the Democratic margin in an election on the Democratic margin in the following election. We found a local average treatment effect of 17.3 percentage points. Our results further support the literature claiming a large incumbency effect, though we show its strength may be decreasing."
format: pdf
bibliography: 530project_bibtex.bib
toc: true
number-sections: true
execute:
  echo: false
  message: false
  warning: false
  results: false
---

@@

Incumbents in the U.S. House of Representatives are reelected at astonishing rates. Since 1964, the reelection rate has never dropped below 85% [@OS1], while the approval rating of the House as a whole has generally remained around 40%, dipping to 20% in recent years [@Gallup1]. Thus, many political scientists spend their careers researching the advantages of incumbency [@Gelman1990].

Many experts have criticized the apparently strong effects of incumbency. Congressmen can use government funds to visit areas of their district, meet with important political players, and send mail to constituents touting their achievements, all of which may indirectly benefit their reelection campaign [@NYT1]. Challengers enjoy none of these benefits.

Previous research has found strong effects of incumbency. Candidates running for reelection routinely raise far more money [@OS2]. Further, they may be able to deter challengers simply by being an incumbent [@Assessing2015]. Overall, Cummings found that in similar districts, incumbent congressmen are more likely to win elections than non-incumbent congressmen from the same party [@Cummings1966].

@@

In practice, it is impossible to determine if the continuity assumption holds for a particular experimental design unless the RDD practitioner has full knowledge of the conditional population densities of the covariates $\mathbf{X}$ with respect to the assigning variable $W$, i.e., $f(\mathbf{X}|W)$, for all covariates that are relevant to $Y$. In empirical RDD, the continuity assumption can be tested with the observed data by (1) testing for the lack of discontinuity in the density of the assignment variable $W$ at threshold $W=w_0$ (McCrary density test) and (2) testing for discontinuity in the first moment of the covariates $\mathbf{X}$ conditional on $W$ at $W=w_0$ [@McCrary2008]. This second test is usually performed by using local linear regressions in the proximity of the line $W=w_0$. In less rigorous language, the first test is meant to determine that the odds of treatment assignment of the barely-controls and the barely-treated are purely random in the proximity of the cutoff, while the second test is to determine if the pretreatment characteristics of barely-controls and the barely-treated are indeed very similar.

@@

The data used in this study came from two primary data sets housed by the Harvard Dataverse and the MIT Election Data and Science Lab. The first data set, _U.S. House 1976–2020 Elections_, is a collection of results of US Congressional House elections between 1976 and 2020 for all congressional districts [@dataset1]. The primary variables used in this data set were _candidate_, _candidate party_, _candidate votes_, and _total votes for election_. The second data set, _Historical Congressional Legislation and District Demographics 1972-2014_, contains a subset of the pretreatment covariates per congressional district we used for this study, namely, _unemployment rate_, _race_, and _level of education_ [@dataset2].

@@

First, we test the continuity of the Democratic vote margin for all $t-1$ elections in our data set. The results of the McCrary density test on the Democratic vote margin at the zero margin cutoff indicates a continuous density on both sides of the cutoff value ($\alpha = 0.05$). A plot of the McCrary density test’s density estimation is provided in @density-all.

```{r}
#| fig-cap: McCrary density test and estimation for all $t-1$ Democratic vote margins.
#| label: density-all

@@

With these three pretreatment covariates chosen, we estimate the discontinuities in their conditional densities conditional on the $t-1$ Democratic election margin at the zero margin value using local linear regression and find that there are no statistically significant discontinuities in either three covariates’ densities. This result can also be expected just by visual inspection of the covariate value versus the $t-1$ Democratic election margin, as in @covars.

```{r}
#| fig-cap: Three pretreatment covariates versus $t-1$ Democratic vote margin.
#| label: covars

@@

```{r}
#| label: basic-rdd
#| fig-cap: RDD analysis of 4,800 elections from 1976 through 2020. The local average treatment effect is 0.173.

@@

We found a clear discontinuity from a slight Democratic loss to a slight Democratic victory. A district that votes barely against the Democrat in election $t-1$ tends to shift further away from Democrats in election $t$, while districts that only marginally elect a Democrat in election $t-1$ move toward Democrats in election $t$. This is a clear effect of the treatment — in this case, electing a Democrat — on the Democratic margin in the next election. See @basic-rdd.

@@

invisible(backwards_rdd(DF))
```

@@

## References
